---
title: "Computational Social Science" 
subtitle: "Data structures"
author: Dr. Thomas Davidson
institute: Rutgers University
date: January 24, 2022
urlcolor: blue
output:
    beamer_presentation:
      theme: "Szeged"
      colortheme: "beaver"
      fonttheme: "structurebold"
      toc: false
      incremental: false
header-includes:
  - \usepackage{hyperref}
  - \usepackage{multicol}
  - \usepackage{caption}
  - \captionsetup[figure]{font=scriptsize}
  - \captionsetup[figure]{labelformat=empty}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(dev = 'pdf')
library("knitr")
library("formatR")

opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
opts_chunk$set(tidy = FALSE)

knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
```


# Plan
- Basic types
- Vectors
- Lists
- Matrices
- Data frames and tibbles


# Basic types
There are four basic types we will be using throughout the class. Here I used them to record some information about one of my cats. In R, it is convention to use the `<-` operator to assign an object to a name.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
# Character (also known as "strings")
name <- "Gary"
# Numeric ("float" in Python)
weight <- 13.2
# Integer ("int" for short)
age <- 4L
# Logical
human <- FALSE
```
\tiny The other two are called `complex` and `raw`. See \href{https://cran.r-project.org/doc/manuals/R-lang.html}{documentation}

# Basic types
There are a few useful commands for inspecting objects.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
print(name) # Prints value in console
class(name) # Shows class of object
typeof(name) # Shows type of object, not always equal to class
```

# Basic types
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
print(weight) # Prints value in console
class(weight) # Shows class of object
typeof(weight) # Shows type of object, not always equal to class
```

# Basic types
We can use the `==` expression to verify the value of an object. We will discuss this in more detail next lecture.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
name == "Tabitha" 
age == 3L
age >= 3L # is greater than
age != 3L # is not
```

# Vectors
A vector is a collection of elements of the *same* type. We can define an empty vector with N elements of a type. Empty vectors assume certain default values depending on the type.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
N <- 5
x <- logical(N)
print(x)
y <- numeric(N)
print(y)
z <- character(N)
print(z)
```

# Vectors
Let's take a closer look at numeric vectors. We can use the combine function `c()` to concatenate multiple values into a vector.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v1 <- c(1,2,3,4,5)
v2 <- c(1,1,1,1,1)
class(v1) # check the class of v1
```

# Vectors
We can easily perform various mathematical operations on numeric vectors
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v1 + v2 # addition
v1 - v2 # subtraction
v1 * v2 # multiplication
sum(v1) # sum over v1
```
\tiny Note how the different methods return different types of outputs. The arithmetic operations return vectors while `sum` returns a numeric value.


# Vectors
What happens if we try to combine objects of different types using combine?
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
t <- c("a", 1, TRUE)
typeof(t)
t
```

# Vectors
There are lots of commands for generating special types of numeric vectors. Note how `N` has already been defined above.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
seq(N) # generates a sequence from 1 to N
rev(seq(N)) # reverses order
rnorm(N) # samples N times from a normal distribution
rbinom(N,1,0.5) # N observations of a single trial with a 0.5 probability of success
```

# Vectors
We can use the help `?` command to find information about each of these commands.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
?rnorm
```

# Vectors
We can use the `index` to access the specific elements of a vector. R uses square brackets for such indexing.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
x <- rnorm(N)
print(x)
print(x[1]) # R indexing starts at 1; Python and some others start at 0
x[1] <- 9 # We can combine indexing with assignment to modify elements
print(x[1])
```

# Vectors
The `head` and `tail` commands are useful when we're working with larger objects. Here we draw 10,000 observations from an normal distribution.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
x <- rnorm(10000)
length(x)
head(x)
tail(x)
head(x, n=10)
```

# Vectors
Retrieve the final element from `x` using indexing.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
# Delete this comment and write answer here
```


# Vectors
Vectors can also contain null elements to indicate missing values, represented by the `NA` symbol.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
x <- c(1,2,3,4,NA)
is.na(x) # The is.na function indicates whether each value is missing.
!is.na(x) # Prepending ! denotes the inverse of a logical operation
```
\tiny `NA` is a logical type but can exist within numeric and character vectors. It is an exception to the rule discussed above regarding the presence of multiple types in the same vector.

# Lists
A list is an object that can contain different types of elements, including basic types and vectors.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
print(v1)
v1.l <- list(v1) # We can easily convert the vector v1 into a list.
print(v1.l)
```

# Lists
Lists have a slightly complex form of indexing. This can be one of the most confusing aspects of R!
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v1.l[1] # The first element of the list contains the vector
v1.l[[1]] # Double brackets allows us to access the vector itself
class(v1.l[1]) # first element is a list
class(v1.l[[1]]) # double indexing gives us the contents
```

# Lists
We can access specific elements of a list by using standard indexing.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v1.l[[1]][1] # Followed by single brackets to access a specific element
v1.l[1][1] # If we're not careful, we will just get the entire sublist
```

# Lists
We can easily combine multiple vectors into a list.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v.list <- list(v1,v2) # We could store both vectors in a list
print(v.list)
v.list[[2]][4] # We can use double brackets to get element 4 of list 1
```

# Lists
We index sublists using double brackets, then specific elements with single brackets.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v.list[[2]][4] # We can use double brackets to get element 4 of list 1
```

# Lists
We can make indexing easier if we start with an empty list and then add elements using a named index via the `$` operator.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v <- list() # initialize empty list
v$v1 <- v1 # the $ sign is used for named indexing
v$v2 <- v2
print(v)
```

# Lists
Combine `$` and square bracket indexing to extract the 5th element of `v1` from `v`.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
# Add code here
```



# Lists
We can define lists more concisely by providing sublists as named arguments.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
cats <- list(names = c("Gary", "Tabitha"), ages = c(4,2))
print(cats)
```
\tiny See \href{https://r4ds.had.co.nz/vectors.html}{Chapter 20} of *R4DS* for more on lists and vectors.

# Matrices
A matrix is a two-dimensional data structure. Like vectors, matrices hold objects of a single type. Here we're defining a matrix using two arguments, the number of rows and columns.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
matrix(nrow=5,ncol=5) # Here there is no content so the matrix is empty
```

# Matrices
We can also pass an argument to define the initial contents of a matrix.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
M <- matrix(0L, nrow=5, ncol=5) # 5x5 matrix of zeros
M
```

# Matrices
We can create a matrix by combining vectors using `cbind`.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
M1 <- cbind(v1,v2) # Treat vectors a columns
print(M1)
```

# Matrices
If we want to treat the vectors as rows, we alternatively use `rbind`. We could also get the same result by *transposing* `M1`.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
M2 <- rbind(v1, v2) # Vectors as rows
print(M2)
print(t(M1)) # t() is the transpose function
```

# Matrices
The `dim` function provides us with information about the dimensions of a given matrix. It returns the number of rows and columns.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
dim(M1) # Shows the dimensions of the matrix
dim(M2)
```

# Matrices
We can get particular values using two-dimensional indexing. By convention `i` denotes the row and `j` the column.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
i <- 1 # row index
j <- 2 # column index
M1[i,j] # Returns element i,j
M1[i,] # Returns row i
M1[,j] # Returns column i
```

# Matrices
Like lists, we can  also name rows and columns to help make indexing easier. The `colnames` and `rownames` functions show the names of each column and row.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
colnames(M1)
rownames(M1)
```

# Matrices
We can use these functions to assign new names.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
colnames(M1) <- c("X", "Y")
rownames(M1) <- seq(1, nrow(M1))
print(M1)
```

# Data frames
*Like its component vectors, a matrix contains data of the same type*. If we have a mix of data types we can use a ``data.frame``. Note how the printed version shows the type of each column.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
df <- as.data.frame(M1) # convert matrix to data frame
class(df)
df$Z <- c("a","b", "c", "d", "e") # assign new column
print(df)
```

# Data frames
We can use indexing in the same way as lists to extract elements. I recommend always using `$` indexing where possible.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
data(iris) # The `data` function loads a built in dataset
head(iris)
iris$Sepal.Length[1] # explicitly call column name
iris[[1]][1] # reference column using index
```

# Tibbles
A `tibble` is the `tidyverse` take on a data.frame. We can easily convert any `data.frame` into a `tibble`.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
library(tidyverse) # the library is required to use the as_tibble function
iris.t <- as_tibble(iris) # convert to tibble
class(iris.t)
```

# Tibbles
Tibbles only show the first ten rows when printing (both look the same in RMarkdown, so we have to use the console to compare.) Tibbles also provide information on the type of each variable.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
#print(iris)
print(iris.t)
```
# Tibbles
Tibbles also tend to provide more warnings when potential issues arise, so they should be less prone to errors than data frames.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize', warning=T }
iris$year # data.frame shows null
iris.t$year # tibble provides a warning
```
  
# Questions?


